\chapter{Fundamentals \textRoman{2}: Linear Optical Response}
\label{cha:fundamentals_b}

\section{Toward electromagnetic response in matter}

% opening paragraph
The previous chapter focused on the ground-state electronic problem.
There, the central question was how electrons arrange themselves under a given external potential.
However, optical properties do not follow from the ground-state alone.
When an electromagnetic field is applied, it drives the electrons away from their static equilibrium and triggers a series of responses.

Therefore, before introducing the dielectric function, we need a language that describes both the applied field and the feedback of matter to that field.
That language is electrodynamics.
The present section does not aim to review a complete lecture of electrodynamics.
Instead, we only extract the ingredients required for the later discussion of the dielectric function, the Kramers--Kronig relations, and the optical properties.
We begin with electromagnetic fields in matter, and then gradually move toward a frequency-dependent response function that can be connected to the electronic structure.

In the present section, we mainly use SI units for the macroscopic electrodynamic equations.
This choice keeps the structure of the Maxwell equations explicit.

\subsection{Maxwell equations in matter}

Let us start by the universal structure of the problem.
When an electromagnetic field enters a medium, it drives charge redistribution and induces currents inside the medium, while these induced sources further modify the field itself.
Therefore, the optical response of matter should be described within the framework of electrodynamics~\cite{griffiths2023introduction}.

At this stage, we do not explore how a specific medium responds.
Instead, we first identify the general field equations that govern electromagnetic dynamics in matter.
They provide the universal backbone of the problem, while the constitutive relations later encode the medium-specific information.

Firstly, we write the macroscopic Maxwell equations for the electric field $\vec{E}(\vec{r},t)$ and magnetic field $\vec{B}(\vec{r},t)$ as:

\begin{equation}
\left\{\begin{aligned}
    \nabla\cdot\vec{E}(\vec{r},t) &= \frac{\rho(\vec{r},t)}{\varepsilon_0} 
    &&\space \text{Gauss's law for electric field} \\
    \nabla\cdot\vec{B}(\vec{r},t) &= 0 
    &&\space \text{Gauss's law for magnetic field} \\
    \nabla\times\vec{E}(\vec{r},t) &= -\frac{\partial}{\partial t}\vec{B}(\vec{r},t) 
    &&\space \text{Faraday's law} \\
    \nabla\times\vec{B}(\vec{r},t) &= \mu_0\vec{J}(\vec{r},t)+\mu_0\varepsilon_0\frac{\partial}{\partial t}\vec{E}(\vec{r},t)
    &&\space \text{Ampère--Maxwell law}
\end{aligned}\right.
\label{maxwell_equations_with_total_sources}
\end{equation}

here, $\rho(\vec{r},t)$ and $\vec{J}(\vec{r},t)$ denote the total charge density and total current density, respectively.

These equations already imply a local conservation law.
More specifically, we take the divergence of the Ampère--Maxwell law,
and use Gauss's law for the electric field to obtain the continuity equation for charge at the macroscopic level:

\begin{equation}
    \frac{\partial}{\partial t}\rho(\vec{r},t)+\nabla\cdot\vec{J}(\vec{r},t)=0
\label{macroscopic_charge_continuity_equation}
\end{equation}

here, the conservation principle encountered in the previous chapter reappears in electrodynamic language.

Secondly, for media, it proves useful to decompose the sources into free and bound contributions:

\begin{equation}
\begin{aligned}
    \rho(\vec{r},t) &= \rho_\mathrm{free}(\vec{r},t)+\rho_\mathrm{bound}(\vec{r},t) \\
    \vec{J}(\vec{r},t) &= \vec{J}_\mathrm{free}(\vec{r},t)+\vec{J}_\mathrm{bound}(\vec{r},t)
\end{aligned}
\label{free_and_bound_source_decomposition}
\end{equation}

here, the free part describes externally supplied or mobile macroscopic sources,
while the bound part describes the response generated internally by the medium.

For discussing the bound electric response, we introduce the polarization field $\vec{P}(\vec{r},t)$ as:

\begin{equation}
    \rho_\mathrm{bound}(\vec{r},t)\coloneqq -\nabla\cdot\vec{P}(\vec{r},t)
\label{bound_charge_from_polarization}
\end{equation}

Accordingly, we define the electric displacement field as:

\begin{equation}
    \vec{D}(\vec{r},t)\coloneqq \varepsilon_0\vec{E}(\vec{r},t)+\vec{P}(\vec{r},t)
\label{electric_displacement_field_definition}
\end{equation}

For the nonmagnetic media considered in the present work, the electric response is the central object.
Therefore, it is sufficient here to keep the Maxwell equations in the form:

\begin{equation}
\left\{\begin{aligned}
    \nabla\cdot\vec{D}(\vec{r},t) &= \rho_\mathrm{free}(\vec{r},t) 
    &&\space \text{Gauss's law for electric displacement} \\
    \nabla\cdot\vec{B}(\vec{r},t) &= 0 
    &&\space \text{Gauss's law for magnetic field} \\
    \nabla\times\vec{E}(\vec{r},t) &= -\frac{\partial}{\partial t}\vec{B}(\vec{r},t) 
    &&\space \text{Faraday's law} \\
    \frac{1}{\mu_0}\nabla\times\vec{B}(\vec{r},t) &= \vec{J}_\mathrm{free}(\vec{r},t)+\frac{\partial}{\partial t}\vec{D}(\vec{r},t)
    &&\space \text{Amp\`ere--Maxwell law in matter}
\end{aligned}\right.
\label{maxwell_equations_in_matter}
\end{equation}

Consequently, we can now characterize the structure of the problem.
The Maxwell equations themselves are universal,
while the medium dependence enters through the induced polarization $\vec{P}(\vec{r},t)$, and therefore through the electric displacement field $\vec{D}(\vec{r},t)$.

In many optical problems, one considers the response to an external electromagnetic perturbation in the absence of free charges and free currents inside the medium.
Under this condition, equation~\eqref{maxwell_equations_in_matter} shows that the central issue is no longer the field equations themselves,
but how the medium develops polarization under the applied electric field.
Once the relation between $\vec{P}(\vec{r},t)$ and $\vec{E}(\vec{r},t)$ is specified, the electromagnetic response becomes well defined.
This is precisely the point where macroscopic electrodynamics meets microscopic electronic structure,
and it also provides the starting point for introducing the dielectric function\cite{bernasconi2011first}.

\subsection{Polarization, electric displacement, and constitutive relations}

We have established the universal field equations in the previous subsection.
However, those equations do not explore how matter responds to the applied field.
To close the problem, we still need a relation that connects the induced polarization to the electric field.
This relation is the constitutive relation, and it is here that the medium-specific information enters~\cite{lima2022usefulness}.

Here, the central object is the polarization field $\vec{P}(\vec{r},t)$.
It describes how the medium rearranges its internal charges under the applied electric field.
Once $\vec{P}(\vec{r},t)$ is known, equation~\eqref{electric_displacement_field_definition} immediately gives the electric displacement field $\vec{D}(\vec{r},t)$.
Therefore, instead of treating $\vec{D}(\vec{r},t)$ as an independent quantity, we may regard it as a convenient way to encode the electric response of matter.

Firstly, we start from the most general linear-response form.
For any time $t$, the polarization may depend not only on the electric field at the same point,
but also on fields acting earlier and in a surrounding spatial region.
Therefore, before introducing the simplified form used in the present work,
we write the constitutive relation in its general nonlocal form in space and time~\cite{royo2019first}:

\begin{equation}
    P_\alpha(\vec{r},t) =
    \varepsilon_0\sum_\beta\int\dif^3{r'}
    \int_{-\infty}^{t}\dif{t'}
    \,\chi_{\alpha\beta}(\vec{r},t;\vec{r}\,',t')\,E_\beta(\vec{r}\,',t')
\label{general_nonlocal_constitutive_relation}
\end{equation}

here, $\chi_{\alpha\beta}(\vec{r},t;\vec{r}\,',t')$ describes the contribution of the electric field acting at $(\vec{r}',t')$ to the polarization that develops later at $(\vec{r},t)$.
In this sense, the medium does not have to respond only at one point and one instant.
Instead, it may spread the response over space and retain memory in time.
At the same time, the upper limit $t$ in equation~\eqref{general_nonlocal_constitutive_relation} keeps the causal order explicit: the perturbation acts first, and the response follows.

For the present work, we focus on the linear response in the long-wavelength regime.
Under this approximation, spatial nonlocality drops out, and the constitutive relation retains only the memory of the earlier perturbation.
Therefore, it reduces to the time-domain constitutive relation:

\begin{equation}
    P_\alpha(\vec{r},t)=
    \varepsilon_0\sum_\beta\int_{-\infty}^{t}\dif{t'}\,\chi_{\alpha\beta}(t-t')\,E_\beta(\vec{r},t')
\label{time_domain_constitutive_relation}
\end{equation}

here, $\chi_{\alpha\beta}(t-t')$ denotes the time-dependent susceptibility tensor.
This expression demonstrates that the polarization does not follow the electric field instantaneously.
Instead, the medium retains a memory of the earlier perturbation.

Optical experiments usually probe the response of a medium under an oscillating electromagnetic field.
Therefore, once we enter the linear-response regime, the frequency domain provides the natural language for discussing the optical response~\cite{han2023temperature}.
Next, we transform equation~\eqref{time_domain_constitutive_relation} to the frequency domain, and obtain:

\begin{equation}
    P_\alpha(\vec{r},\omega)=\varepsilon_0\sum_\beta\chi_{\alpha\beta}(\omega)\,E_\beta(\vec{r},\omega)
\label{frequency_domain_polarization_susceptibility_relation}
\end{equation}

here, $\chi_{\alpha\beta}(\omega)$ denotes the frequency-dependent electric susceptibility tensor.
This equation already shows that the susceptibility determines the linear electric response of the medium.

Substituting equation~\eqref{frequency_domain_polarization_susceptibility_relation} into equation~\eqref{electric_displacement_field_definition},
we then express the electric displacement field in terms of the electric susceptibility tensor:

\begin{equation}
\begin{aligned}
    D_\alpha(\vec{r},\omega)
    &=\varepsilon_0 E_\alpha(\vec{r},\omega)+P_\alpha(\vec{r},\omega) \\
    &=\varepsilon_0\sum_\beta\left[\delta_{\alpha\beta}+\chi_{\alpha\beta}(\omega)\right]E_\beta(\vec{r},\omega)
\end{aligned}
\label{electric_displacement_from_susceptibility}
\end{equation}

It now shows that the combination $\delta_{\alpha\beta}+\chi_{\alpha\beta}(\omega)$ naturally enters the linear electric response.

This observation directly motivates the definition of the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$:

\begin{equation}
    \varepsilon_{\alpha\beta}(\omega)
    \coloneqq\delta_{\alpha\beta}+\chi_{\alpha\beta}(\omega)
\label{dielectric_tensor_definition}
\end{equation}

Then, the frequency-domain constitutive relation takes the compact form:

\begin{equation}
    D_\alpha(\vec{r},\omega)
    = \varepsilon_0\sum_\beta\varepsilon_{\alpha\beta}(\omega)E_\beta(\vec{r},\omega)
\label{frequency_domain_constitutive_relation_with_dielectric_tensor}
\end{equation}

The dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ thus provides a compact description of the linear electric response.
Starting from the polarization field $\vec{P}(\vec{r},t)$, we introduce the electric susceptibility tensor $\chi_{\alpha\beta}(\omega)$ and then arrive at the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$.
With this preparation, we next discuss the anisotropy of the dielectric tensor.

\subsection{Anisotropy and the dielectric tensor}

The previous subsection presented the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ as a compact description of the linear electric response.
We now examine how symmetry constrains its form~\cite{capper2019optical}.
When the medium responds identically along all directions, the dielectric tensor reduces to a scalar form.
If the response changes with direction, the full tensor form must be retained.

Firstly, for an isotropic medium,
the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ reduces to a scalar dielectric function $\varepsilon(\omega)$:

\begin{equation}
    \varepsilon_{\alpha\beta}(\omega)
    = \varepsilon(\omega)\delta_{\alpha\beta}
\label{isotropic_dielectric_tensor}
\end{equation}

here, the scalar dielectric function $\varepsilon(\omega)$ already determines the full linear electric response,
since the isotropic medium does not distinguish one Cartesian direction from another.
The tensor structure, therefore, no longer carries additional directional information.

Secondly, crystalline media usually distinguish directions.
In that case, anisotropy survives, and we must retain the full tensor form.
When we choose the Cartesian axes along the principal directions of the crystal,
the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ takes the principal-axis form:

\begin{equation}
    \varepsilon(\omega) = \begin{pmatrix}
        \varepsilon_{xx}(\omega) & 0 & 0 \\
        0 & \varepsilon_{yy}(\omega) & 0 \\
        0 & 0 & \varepsilon_{zz}(\omega)
    \end{pmatrix}
\label{principal_axis_dielectric_tensor}
\end{equation}

here, each diagonal component describes the dielectric response along one principal axis,
and anisotropy enters through the difference among these components.

For the present work, this directional dependence plays a central role~\cite{slavich2024exploring}.
We later compare the optical response within the structural plane and perpendicular to that plane.
Therefore, it is convenient to introduce the in-plane dielectric function $\varepsilon_{\varparallel}(\omega)$ and the out-of-plane dielectric function $\varepsilon_{\perp}(\omega)$:

\begin{equation}
    \varepsilon(\omega) = \begin{pmatrix}
        \varepsilon_{\varparallel}(\omega) & 0 & 0 \\
        0 & \varepsilon_{\varparallel}(\omega) & 0 \\
        0 & 0 & \varepsilon_{\perp}(\omega)
    \end{pmatrix}
\label{inplane_outofplane_dielectric_tensor}
\end{equation}

here, the in-plane dielectric function $\varepsilon_{\varparallel}(\omega)$ describes the response to an electric field within the structural plane,
while the out-of-plane dielectric function $\varepsilon_{\perp}(\omega)$ describes the response to an electric field normal to that plane.
This notation directly prepares the comparison between the in-plane and out-of-plane dielectric spectra and optical properties~\cite{guo2021complete}.

This anisotropy also propagates to other optical quantities derived from the dielectric tensor~\cite{sujith2023large}.
Once the dielectric response changes with direction, the refractive index, the extinction coefficient, the reflectivity, and the energy-loss spectrum also inherit that directional dependence.
Therefore, the tensor structure does not only refine the notation.
It also determines how we organize and interpret the later optical results.

Starting from the polarization field $\vec{P}(\vec{r},t)$, we introduced the electric susceptibility tensor $\chi_{\alpha\beta}(\omega)$ and then arrived at the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$.
The dielectric function, therefore, emerges naturally from the constitutive relation that closes Maxwell electrodynamics in matter.
At the same time, anisotropy makes the directional dependence of this response explicit.
With this preparation, we next introduce the dielectric function in a more explicit form.

\newpage
\section{From electronic response to the dielectric function}

% opening paragraph
The previous section established the electrodynamic framework of optical response.
There, we introduced the polarization field $\vec{P}(\vec{r},t)$,
the electric susceptibility tensor $\chi_{\alpha\beta}(\omega)$,
and the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$.
These quantities complete Maxwell electrodynamics in matter.
We now recover the response function that electronic structure calculations actually provide,
which is the frequency-dependent dielectric function~\cite{onida2002electronic}.

In this section, we start by expressing the polarization response in the language of electric susceptibility and the complex dielectric function.
We then identify the microscopic electronic processes that generate its imaginary part.
Next, we formulate this response for anisotropic systems,
since the present work compares the in-plane and out-of-plane optical channels explicitly.
Therefore, we advance from the general constitutive picture to a response function that can be calculated, interpreted, and subsequently converted into optical observables.

\subsection{From polarization response to electric susceptibility and complex dielectric function}

The previous section established the constitutive chain from the polarization field $\vec{P}(\vec{r},t)$ to the electric susceptibility tensor $\chi_{\alpha\beta}(\omega)$ 
and the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ through equations~\eqref{frequency_domain_polarization_susceptibility_relation} and~\eqref{dielectric_tensor_definition}.

We now turn to the dielectric function, which serves as the central quantity in optical response.
A time-dependent electric field drives the electrons away from equilibrium and induces a polarization response.
In linear optics, incident light provides this driving field.
We describe this response in terms of the electric susceptibility and the dielectric function.

The electric susceptibility tensor $\chi_{\alpha\beta}(\omega)$ measures how strongly the polarization component $P_\alpha$ responds to the field component $E_\beta$, while the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ collects this response into the form most directly used in optical analysis~\cite{gajdovs2006linear}.

At this stage, we have clarified that the dielectric function describes how internal charges polarize in response to an external electric field and how that response varies with frequency.
When we enter the frequency domain, the optical response no longer appears as a static redistribution of charge.
Instead, the field drives the electrons continuously, and the medium answers with a frequency-dependent polarization.
For this reason, the dielectric function becomes the central response quantity for discussing optical performance.

Under a monochromatic electric field, part of the polarization stays in phase with the field,
while another part lags behind it.
The first part stores energy in the medium, and the second part transfers energy from the field into the medium.
We therefore write the dielectric tensor in complex form as:

\begin{equation}
    \varepsilon_{\alpha\beta}(\omega)
    = \varepsilon_{1,\alpha\beta}(\omega)
    + \mathrm{i}\,\varepsilon_{2,\alpha\beta}(\omega)
\label{complex_dielectric_tensor_definition}
\end{equation}

here, $\varepsilon_{1,\alpha\beta}(\omega)$ denotes the real part of the dielectric tensor,
and $\varepsilon_{2,\alpha\beta}(\omega)$ denotes the imaginary part of the dielectric tensor.

The real part denotes the dispersive part of the response.
It describes how the medium polarizes with the field and therefore how it stores electromagnetic energy.

The imaginary part denotes the absorptive part of the response.
It describes how the field drives electronic excitations and therefore how it transfers energy into the medium~\cite{ambrosch2006linear}.

For the later discussion on optical observables,
when the field polarization is fixed or working along a principal direction of the dielectric tensor,
we suppress the tensor indices and write the dielectric response in scalar form.

Therefore, we can describe the corresponding complex dielectric function as a degenerate expression:

\begin{equation}
    \varepsilon(\omega)
    = \varepsilon_1(\omega)
    + \mathrm{i}\,\varepsilon_2(\omega)
\label{complex_dielectric_function_definition}
\end{equation}

here, $\varepsilon_1(\omega)$ denotes the real part of the dielectric function,
and $\varepsilon_2(\omega)$ denotes the imaginary part of the dielectric function.

This scalar form represents the dielectric response along a specific direction.
Here, the complex dielectric function describes the response of medium to an external electric field.
we can therefore derive 
the absorption coefficient $\alpha(\omega)$,
the refractive index $n(\omega)$,
the extinction coefficient $k(\omega)$,
the reflectivity $R(\omega)$,
and the energy-loss function $L(\omega)$.

It represents a unified framework for the linear optical properties~\cite{djurivsic2002progress}.

These two parts of the dielectric function together determine the optical response.
In practical calculations, the imaginary part $\varepsilon_2(\omega)$ comes from electronic excitations,
whereas the real part $\varepsilon_1(\omega)$ follows from causality.

For the xene systems considered in the present work,
we accordingly resolve the dielectric response for the in-plane and out-of-plane channels for separate analysis.
In anisotropic xenes, the response along the structural plane usually differs substantially from that along the perpendicular direction~\cite{marinopoulos2002anisotropy}.

At this stage, we arrive at the complex dielectric function as the central response quantity of the present chapter.
Here, microscopic electronic processes generate the imaginary part $\varepsilon_2(\omega)$,
and the Kramers--Kronig relation then determines the real part $\varepsilon_1(\omega)$~\cite{persson2001first}.
We now turn to the microscopic origin of the imaginary part $\varepsilon_2(\omega)$.

\subsection{Microscopic origin of the imaginary part}

For the complex dielectric function,
the previous subsection discussed that the imaginary part $\varepsilon_2(\omega)$ characterizes the absorptive response of the medium.
We now connect this quantity to the microscopic electronic transitions in the crystal.

When a monochromatic electric field acts on a crystal,
it drives electrons from occupied states to empty states,
and therefore induces interband transitions.
In this process, the external field transfers the energy $\hbar\omega$ to the electronic system,
and the medium absorbs this energy through a dissipative response~\cite{toudert2017interband}.
To describe this dissipative response quantitatively,
we now introduce the optical conductivity tensor $\sigma_{\alpha\beta}(\omega)$ within linear response theory~\cite{kubo1957statistical}.

Firstly, we express the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ through the optical conductivity tensor $\sigma_{\alpha\beta}(\omega)$ as:

\begin{equation}
    \varepsilon_{\alpha\beta}(\omega)
    = \delta_{\alpha\beta} + \frac{\mathrm{i}}{\varepsilon_0\omega}\,
    \sigma_{\alpha\beta}(\omega)
\label{dielectric_tensor_from_optical_conductivity}
\end{equation}

Once we decompose this optical conductivity tensor into its real part $\sigma_{1,\alpha\beta}(\omega)$ and imaginary part $\sigma_{2,\alpha\beta}(\omega)$, 
the previous equation immediately gives:

\begin{equation}
    \varepsilon_{2,\alpha\beta}(\omega)
    = \frac{\sigma_{1,\alpha\beta}(\omega)}{\varepsilon_0\omega}
\label{imaginary_dielectric_from_real_conductivity}
\end{equation}

This relation demonstrates that the imaginary part of the dielectric response directly describes the energy-absorbing part of the induced current,
which leads to dissipation under the external field.

Next, we describe this microscopic process in a periodic solid as an interband transition from an occupied valence state $\ket{\psi_{v\vec{k}}}$ to an empty conduction state $\ket{\psi_{c\vec{k}}}$.
Here, $\vec{k}$ denotes the crystal momentum in the first Brillouin zone.
In the optical regime, the photon momentum is negligible on the scale of the Brillouin zone.

In the optical regime, the photon momentum remains negligible on the scale of the Brillouin zone,
and the transition therefore occurs nearly vertically in $\vec{k}$-space~\cite{fox2010optical}.

Therefore, in $\vec{k}$-space, the transition proceeds effectively vertically.
The electric field drives electronic motion directly.
Accordingly, the transition strength depends on the velocity operator $\hat{v}_\alpha$ rather than on the band energies alone~\cite{adler1962quantum, wiser1963dielectric}.

In the present work, we adopt the independent-particle approximation (IPA).
The imaginary part of the dielectric tensor $\varepsilon_{2,\alpha\beta}(\omega)$ takes the standard interband form~\cite{d2016accurate}:

\begin{equation}
    \varepsilon_{2,\alpha\beta}(\omega)\propto
    \sum_{v,c}
    \int_{\mathrm{BZ}}\frac{\dif^3{k}}{(2\pi)^3}\,
    \bra{\psi_{c\vec{k}}}\hat{v}_{\alpha}\ket{\psi_{v\vec{k}}}
    \bra{\psi_{v\vec{k}}}\hat{v}_{\beta}\ket{\psi_{c\vec{k}}}
    \delta\left(E_{c\vec{k}} - E_{v\vec{k}} - \hbar\omega\right)
\label{independent_particle_imaginary_dielectric_tensor}
\end{equation}

here, the valence-band index $v$ and conduction-band index $c$ label the occupied and empty bands, respectively.
The band energies $E_{v\vec{k}}$ and $E_{c\vec{k}}$ describe the initial and final states.
The velocity operator $\hat{v}_{\alpha}$ determines the transition strength along the $\alpha$ direction.
Meanwhile, the Dirac delta function $\delta\left(E_{c\vec{k}}-E_{v\vec{k}}-\hbar\omega\right)$ enforces the energy-conservation condition:

\begin{equation}
    E_{c\vec{k}}-E_{v\vec{k}}=\hbar\omega
\label{dirac_energy_conservation_condition}
\end{equation}

Equation~\eqref{independent_particle_imaginary_dielectric_tensor} describes the microscopic origin of the absorptive response.
For a transition to contribute to the absorptive response,
the initial state must be occupied,
the final state must be empty, the transition energy
$E_{c\vec{k}}-E_{v\vec{k}}$ must match the photon energy $\hbar\omega$,
and the velocity matrix elements
$\bra{\psi_{c\vec{k}}}\hat{v}_{\alpha}\ket{\psi_{v\vec{k}}}$
and $\bra{\psi_{v\vec{k}}}\hat{v}_{\beta}\ket{\psi_{c\vec{k}}}$
must remain finite.
These finite matrix elements define the optical selection rule and identify the optically allowed transitions.
Therefore, when many optically allowed transitions share the same excitation energy and show strong transition strength,
a peak appears in $\varepsilon_{2,\alpha\beta}(\omega)$.

This point clarifies what the calculated spectra describe.
The imaginary part $\varepsilon_{2,\alpha\beta}(\omega)$ reflects the band gap,
the band dispersion, the occupation pattern,
and the direction-resolved transition matrix elements.
At this stage, $\varepsilon_{2,\alpha\beta}(\omega)$ provides a direct bridge between the electronic structure and the optical absorption channel.

In the present work, we further adopt the long-wavelength limit.
Accordingly, we evaluate the dielectric response within the independent-particle approximation built from the ground-state electronic structure.
Under this approximation, the calculated $\varepsilon_{2,\alpha\beta}(\omega)$ describes the single-particle interband absorption,
and it identifies the main transition channels in the optical spectra.
Accordingly, it captures the principal onset energies, the major absorption peaks,
and the directional contrast relevant to the systems studied here.

The central point is that the tensor indices $\alpha$ and $\beta$ enter directly through the velocity matrix elements.
Therefore, the microscopic transition strength already resolves the field polarization and the crystal directions.
Once the crystal responds differently along different directions,
the imaginary part $\varepsilon_{2,\alpha\beta}(\omega)$ already exhibits this anisotropy at the dielectric-response level.
We thus arrive at a microscopic description of $\varepsilon_{2,\alpha\beta}(\omega)$,
which collects the energy-conserving interband transitions and weights them by their directional dependence.
This result naturally leads us to the dielectric response in anisotropic media.~\cite{huang2023optical}.

\subsection{Dielectric response in anisotropic media}

The previous subsection established the microscopic expression for the imaginary part $\varepsilon_{2,\alpha\beta}(\omega)$.
We now describe the dielectric response of anisotropic media in a direction-resolved form~\cite{slavich2024exploring}.

For an anisotropic medium,
each principal direction corresponds to a specific complex dielectric response.

Once we determine the principal directions,
equation~\eqref{principal_axis_dielectric_tensor} describes the principal-axis form of the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$.
For each principal component with $\lambda=x,y,z$, 
we write the complex dielectric function in a direction-resolved form:

\begin{equation}
    \varepsilon_{\lambda\lambda}(\omega) =
    \varepsilon_{1,\lambda\lambda}(\omega)
    + \mathrm{i}\,\varepsilon_{2,\lambda\lambda}(\omega)
\label{direction_resolved_complex_dielectric_function}
\end{equation}

here, the real part $\varepsilon_{1,\lambda\lambda}(\omega)$ describes the directional dispersive response,
whereas the imaginary part $\varepsilon_{2,\lambda\lambda}(\omega)$ describes the directional absorptive response.
As different principal components produce different optical responses along different directions, 
optical anisotropy appears~\cite{nye1985physical}.

For xenes, the dielectric response separates into the structural plane and the normal direction~\cite{matthes2016influence}.

In the present work,
we denote the dielectric response within the structural plane by the in-plane dielectric function $\varepsilon_{\varparallel}(\omega)$,
whereas we denote the response normal to that plane by
the out-of-plane dielectric function $\varepsilon_{\perp}(\omega)$.
This notation keeps the later comparison clear.

Once we decompose the dielectric response into directional channels,
the derived optical quantities inherit the same anisotropy~\cite{alonso2004optical}.

At this stage, we have established the direction-resolved complex dielectric response for anisotropic media.

For each principal direction, the real and imaginary parts are not independent.
In the next section,
we introduce causality as the fundamental constraint on the linear response,
and then derive the Kramers--Kronig relations for each directional component.

\newpage
\section{Causality and the Kramers--Kronig relations}

% opening paragraph
The previous section established a direction-resolved complex dielectric response for anisotropic media.
We now identify the constraint that connects the real and imaginary parts of each directional component.

Here, the major physical constraint is causality~\cite{toll1956causality}.
Specifically, causality means that the polarization response cannot occur before the external electric field is applied.
Equivalently, in the time domain, the response kernel vanishes when the response time precedes the driving time.
This temporal order determines the analytic structure of the response function in frequency space.
As a result, for each complex component,
the real and imaginary parts cannot vary independently.
Instead, they are related through the Kramers--Kronig relations~\cite{lucarini2005kramers}.

Therefore, for the dielectric response, once we calculate the absorptive part,
we can determine the dispersive part and construct the complete dielectric response.
This connection completes the dielectric response and opens the derivation of the optical properties.

\subsection{Causality and analytic structure of response functions}

Firstly, to express causality explicitly, 
we start from the time-domain linear relation between
the polarization field $P_{\alpha}(\vec{r},t)$ and
the electric field $E_{\beta}(\vec{r}\,',t')$~\cite{dressel2002electrodynamics}:

\begin{equation}
    P_{\alpha}(\vec{r},t) = 
    \varepsilon_{0}\sum_{\beta}\int_{-\infty}^{\infty}\dif{t'}
    \int\dif^3{r'}\,\chi_{\alpha\beta}(\vec{r},\vec{r}\,',t-t')
    E_{\beta}(\vec{r}\,',t')
\label{general_linear_response_polarization}
\end{equation}

here, $\chi_{\alpha\beta}(\vec{r},\vec{r}\,',t-t')$ denotes the time-domain electric susceptibility tensor.
This equation characterizes the polarization response at position $\vec{r}$ and time $t$ to an electric field applied at position $\vec{r}\,'$ and time $t'$.

Therefore, causality requires that the medium responds only to earlier external driving fields~\cite{stefanski2022analytical}:

\begin{equation}
    \chi_{\alpha\beta}(\vec{r},\vec{r}\,',t-t')=0
    \qquad \text{for $t<t'$}
\label{causal_condition_susceptibility_kernel}
\end{equation}

This condition defines a retarded response kernel~\cite{landau2013electrodynamics}.
It excludes any polarization response before the external electric field acts.

In the present work,
we consider the optical long-wavelength limit,
where the spatial nonlocality of the dielectric response becomes negligible~\cite{agranovich2013crystal}.
Afterwards, the linear response then reduces to a local frequency-dependent tensor relation~\cite{franta2023dispersion}:

\begin{equation}
    P_{\alpha}(\omega) =
    \varepsilon_{0}\sum_{\beta}\chi_{\alpha\beta}(\omega)
    \,E_{\beta}(\omega)
\label{frequency_domain_polarization_response}
\end{equation}

here, $\chi_{\alpha\beta}(\omega)$ denotes the frequency-domain electric susceptibility tensor.
It characterizes the response amplitude for each Cartesian component at angular frequency $\omega$.
Meanwhile, it also describes the phase delay of the response at the same frequency~\cite{koutserimpas2024time}.

We then express the dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ via the electric susceptibility tensor $\chi_{\alpha\beta}(\omega)$~\cite{franta2020symmetry}:

\begin{equation}
    \varepsilon_{\alpha\beta}(\omega) =
    \delta_{\alpha\beta}
    + \chi_{\alpha\beta}(\omega)
\label{dielectric_tensor_from_susceptibility}
\end{equation}

Subsequently, the causal condition in equation~\eqref{causal_condition_susceptibility_kernel}
gives the frequency-domain susceptibility in the form:

\begin{equation}
\chi_{\alpha\beta}(\omega) =
    \int_{0}^{\infty}
    \chi_{\alpha\beta}(t)
    \,\mathrm{e}^{\,\mathrm{i}\omega t}
    \dif{t}
\label{frequency_domain_susceptibility_from_causality}
\end{equation}

This representation reveals the analytic structure directly.
Since the time integral starts from $t=0$, the response function remains analytic in the upper half of the complex frequency plane~\cite{nussenzveig1972causality}.

For a real time-domain response kernel, the susceptibility $\chi_{\alpha\beta}(\omega)$
and dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$ further satisfy the following condition:

\begin{equation}
\left\{\begin{aligned}
    \chi_{\alpha\beta}(-\omega)
    &= \chi_{\alpha\beta}^{*}(\omega) \\
    \varepsilon_{\alpha\beta}(-\omega)
    &= \varepsilon_{\alpha\beta}^{*}(\omega)
\end{aligned}\right.
\label{reality_condition_response_functions}
\end{equation}

Therefore, in the principal-axis representation,
for each complex dielectric component $\varepsilon_{\lambda\lambda}(\omega)$,
the real part $\varepsilon_{1,\lambda\lambda}(\omega)$ is an even function of frequency,
whereas the imaginary part $\varepsilon_{2,\lambda\lambda}(\omega)$ is an odd function:

\begin{equation}
\left\{\begin{aligned}
    \varepsilon_{1,\lambda\lambda}(-\omega)
    &= \varepsilon_{1,\lambda\lambda}(\omega) \\
    \varepsilon_{2,\lambda\lambda}(-\omega)
    &= -\varepsilon_{2,\lambda\lambda}(\omega)
\end{aligned}\right.
\label{parity_of_dielectric_components}
\end{equation}

At this stage, the previous discussion has established the analytic structure of the dielectric response in frequency space~\cite{leon2026momentum}.
We next derive the Kramers--Kronig relations between its real and imaginary parts.

\subsection{Kramers--Kronig relations for the dielectric function}

Draft - Draft - Draft - Draft - Draft - Draft % Draft

\subsection{Static limit and physical interpretation}

\newpage
\section{From the dielectric function to optical properties}

% opening paragraph

\subsection{Complex refractive index, refractive index, and extinction coefficient}

\subsection{Absorption coefficient}

\subsection{Reflectivity}

\subsection{Energy-loss function}
