\chapter[Functional Properties of Penta-Bipyramid Boron Allotropes]{Superconductivity, Magnetism, and Directional Plasmonic Modes in
Two-Dimensional Penta-Bipyramid Boron Allotropes}
\label{cha:project_2}

\section{Abstract}
\label{2_abstract}

Bulk orthorhombic boron, o-\ce{B14}, represents an intriguing, recently predicted,
three-dimensional (3D) boron allotrope.
Its structure consists of edge-sharing pentagonal bipyramids
with diverse bond lengths.
These structural motifs facilitate the formation of unusual seven-centre-two-electron $\pi$-bonds,
which were predicted for the first time in 3D boron allotropes.
The presence of these bonds highlights the complexity and versatility of boron chemistry under ambient conditions.
Moreover, this material is predicted to be a superconductor with a critical temperature $T_c$ of \qty{29.1}{K}~\cite{han2023superconducting}.

In the present work, we investigate the electronic and optical properties of bulk o-\ce{B14},
its monolayer and bilayer forms,
and their hydrogen-terminated counterparts using first-principles calculations.
These two-dimensional structures are predicted to remain dynamically stable and to exhibit a remarkable variety of electronic behaviour,
including semiconductivity with a band-gap of \qty{0.67}{\electronvolt} in the bilayer,
superconductivity with a $T_c$ of \qty{24.3}{K} in the H-terminated bilayer,
and magnetism in the monolayer.
The directional bonding network drives anisotropic plasmonic and optical responses,
as well as absorption in the visible region of the spectrum.
These properties suggest promising opportunities for applications in advanced optoelectronic, plasmonic and quantum devices.

\clearpage
\section{Introduction}
\label{2_introduction}

Elemental boron possesses a unique combination of light weight, exceptional hardness, and thermal and chemical stability.
Due to these properties, it has potential for high-technology applications such as catalysis, sensors, energy storage, piezoelectrics, and nanodevices~\cite{ranjan2024properties}.
Boron exists in multiple allotropes with complex crystal structures arising from its electron-deficient bonding and small atomic size.
The most well-characterized and stable crystalline forms are $\alpha$-rhombohedral boron ($\alpha$-B, 12 atoms per unit cell),
$\beta$-rhombohedral boron ($\beta$-B, 105--108 atoms per unit cell)~\cite{widom2008symmetry},
and $\gamma$-orthorhombic boron ($\gamma$-B, 28 atoms per unit cell)~\cite{an2016new}.
$\alpha$-B consists of \ce{B12} icosahedral building blocks and forms at low temperature,
whereas $\beta$-B has the same structural motif, but also includes interstitial B atoms.
$\beta$-B forms at room temperature and is the most stable structure.
$\gamma$-B is also composed of \ce{B12} icosahedral units and forms at high pressure and
temperature.
The structures are predominantly characterized by covalent bonding,
but the electron deficiency of boron gives rise to multi-centre bonding,
such as three-centre-two-electron bonds.
Additionally, boron-rich compounds exhibit promising semiconducting and thermoelectric properties,
with potential use in next-generation electronics and energy-harvesting devices.

Generally, the superconducting critical temperature ($T_c$) of elemental superconductors is relatively low at ambient pressure.
A representative elemental superconductor, \ce{Nb}, has a $T_c$ of about \qty{9}{K}.
Under high pressure, the $T_c$ can be significantly increased compared with the ambient-pressure value,
but high pressure hinders experimental studies and practical applications.
Therefore,
it is of high priority to explore elemental superconductors with a high $T_c$ under low or ambient pressure.
Interestingly, superconductivity with a $T_c$ of \qty{35}{K} has been reported in a three-dimensional (3D) kagome-like boron allotrope under \qty{7}{GPa} on the basis of \textit{ab initio} calculations~\cite{ouyang2025superconductivity},
while earlier work predicts a $T_c$ exceeding \qty{15}{K} in 3D boron allotropes~\cite{zhang2020superconducting}.

Boron can also form isolated clusters which exhibit many intriguing properties,
including fluxionality~\cite{pan2019fluxional}.
Fluxionality describes the ability to undergo continuous structural rearrangements without breaking chemical bonds in a way that permanently alters the system.
Recent studies have focused on two-dimensional (2D) forms of boron,
including monolayer borophene, where many different structures have been observed.
Recent review papers have summarized these developments:
Wang \textit{et al.}~\cite{wang2024borophene} have summarized the synthesis and properties of borophene,
while Mishra \textit{et al.}~\cite{mishra2024borophene} have reviewed the potential of borophene in various technological fields such as batteries, supercapacitors, and fuel cells, as well as the associated synthesis techniques.
Other recent reviews of borophene focus on experimental aspects~\cite{mohanty2023review,yang2023borophenes,kumar2024,xie2020two},
and provide a survey of structures starting from the bulk allotropes, boron clusters,
potential building blocks for 2D boron crystals, and 2D few-layered boron sheets~\cite{sun2017two}.
As a light 2D material with good flexibility,
borophene has potential applications in flexible and wearable devices, biomedicines,
and biosensors~\cite{mozvashi2021mechanical,zhang2017elasticity}.

From the theoretical side,
2D boron has also attracted significant recent interest due to its complex electronic structure,
which can give rise to predicted semiconducting behaviour~\cite{qureshi2024bilayer},
including in twisted bilayer borophene~\cite{jin2022tuning},
superconductivity~\cite{gao2017prediction,han2023superconducting},
magnetism~\cite{zhu2019magnetic}, and topological states~\cite{wang2025second,rakshit2025exploring,wang2019serendipity}.
Furthermore, the $\beta_{12}$ sheet has been found to host gapless Dirac cones~\cite{feng2017dirac},
and a bilayer kagome structure has been predicted to host multiple van Hove singularities~\cite{gao2024bilayer}.
Half-metallic ferromagnetism has also been reported in a 2D kagome lattice composed of a triangular arrangement of boron atoms within triangles~\cite{kim2017boron}.

Synthesizing borophene is notably challenging due to several intrinsic and extrinsic factors.
Unlike graphene, borophene lacks a naturally occurring layered bulk counterpart,
which eliminates the possibility of exfoliation-based synthesis.
Its high chemical reactivity,
especially with oxygen and moisture, makes it unstable under ambient conditions,
requiring ultra-high vacuum (UHV) environments for fabrication and handling.
These factors collectively make the reproducible and scalable synthesis of monolayer borophene a significant materials science hurdle.
To overcome the challenges of synthesizing monolayer borophene,
researchers have focused on carefully controlled epitaxial growth techniques such as molecular beam epitaxy (MBE) on inert metal substrates~\cite{chen2022synthesis}.
Another study has achieved growth of $\beta_{12}$- and $\alpha$-sheets of borophene on \ce{Ag}(111) by direct evaporation of a pure boron source in an e-beam evaporator~\cite{zhong2017metastable}.

Substrates stabilize the 2D structure of borophene during growth by promoting specific polymorphs through surface templating.
Maintaining UHV conditions during synthesis helps prevent oxidation and contamination.
Another strategy to enhance stability uses hydrogen termination,
which passivates the highly reactive surface boron atoms.
This passivation can reduce the tendency to oxidize or react with ambient species,
thereby improving chemical stability in air.
Moreover, hydrogen termination can modify the electronic properties,
potentially opening a band-gap or altering metallic behaviour,
which may be advantageous for specific applications~\cite{hou2020ultrastable,li2021synthesis}.
Large-scale synthesis of freestanding atomic sheets of borophene has been reported through a novel liquid-phase exfoliation approach,
where electron microscopy confirmed the presence of the $\beta_{12}$ and $\chi_{3}$ phases,
as well as their intermediate phases~\cite{ranjan2019freestanding}.

Different from flat borophenes,
a novel highly stable 2D boron structure has been recently predicted on the basis of \textit{ab initio} calculations.
This structure is composed of \ce{B12} clusters linked by single B-B bonds and is a narrow band-gap semiconductor~\cite{zhang2023novel}.
The optical properties of 2D allotropes have also been investigated.
It has been reported that the plasmon frequencies reach the near infrared and even visible regions,
making borophene among the first materials with 2D plasmons at such high energies without doping~\cite{huang2017}.

Recently, a bulk orthorhombic boron allotrope, o-\ce{B14},
with interesting properties was predicted~\cite{han2023superconducting}.
Its structure is characterized by edge-sharing pentagonal bipyramids with varying B-B bond lengths ranging from \qtyrange{1.644}{1.966}{\angstrom}.
These structural motifs facilitate the formation of peculiar seven-centre-two-electron $\pi$-bonds,
which were predicted for the first time in 3D boron allotropes.
Remarkably, this material is predicted to be a superconductor with a transition temperature of \qty{29.1}{K}.

This orthorhombic o-\ce{B14} structure is a particularly promising candidate for further investigation,
as it introduces the pentagonal \ce{B7} bipyramidal motif as a fundamental structural unit in crystalline boron,
which differs markedly from the \ce{B12} icosahedral units that dominate most known boron phases.
This unusual topology leads to highly delocalized multicentre bonding,
including the predicted seven-centre-two-electron $\pi$ bonds,
suggesting that the electronic structure may host unconventional features such as anisotropic transport, unusual phonon behaviour,
and potentially strong electron-phonon coupling.
Since the phase is predicted to be energetically competitive and dynamically stable at ambient pressure, \textit{ab initio} studies of
its electronic band structure, phonon spectrum, and superconducting properties would be valuable,
as would studies that explore how the unique bipyramidal building blocks influence the broader physics of electron-deficient boron networks.
Moreover, this also raises the question of whether it may be stable in 2D form,
where the interlayer connectivity consists only of single B-B bonds.

In the present work, we use density-functional theory (DFT) calculations to investigate the electronic and optical properties of bulk o-\ce{B14} and its mono- and bilayers.
Hydrogen termination of the 2D structures is also considered.
Interestingly, bilayer o-\ce{B14} is found to be a semiconductor with a band-gap of around
\qty{0.69}{\electronvolt} as calculated using the HSE06 hybrid functional.
The H-terminated bilayer, however, is predicted to be superconducting with a $T_c$ of \qty{24.3}{K},
while monolayer o-\ce{B14} is found to be magnetic.
The optical properties reveal distinctive anisotropic behaviour and the existence of plasmons in the near- to mid-infrared range for in-plane excitation,
as well as the ultraviolet energy range for out-of-plane excitation,
broadening the potential of boron for applications in optoelectronics, plasmonics, and catalysis.

This chapter is organised as follows:
In Section~\ref{2_calculation_methods} the computational methods are presented,
followed by the results and discussion in Section~\ref{2_results}.
Herein, we describe the electronic, superconducting, and optical properties.
This is followed by the conclusions in Section~\ref{2_conclusions}.

\section{Computation methods}
\label{2_calculation_methods}

The DFT calculations are performed using the Vienna Ab initio Simulation Package (VASP)~\cite{kresse1993ab,kresse1996efficiency,kresse1994ef,kresse1994ab}.
The projector augmented wave (PAW) method pseudopotentials are used to describe the electron-ion interactions~\cite{kresse1999ultrasoft}.
The generalized gradient approximation (GGA) of Perdew--Burke--Ernzerhof (PBE) for the exchange-correlation functional~\cite{PBE-1996} is employed along with the Becke--Johnson damped D3(BJ) dispersion correction to account for long-range van der Waals (vdW) interactions~\cite{D3BJ2005,D3BJ2010,D3BJ2011}.
For bulk o-\ce{B14}, a $\Gamma$-centered $\vec{k}$-point mesh of $34 \times 48 \times 31$ is used with a plane-wave energy cutoff of \qty{600}{\electronvolt}.
The electronic convergence criterion is set to \qty{1e-6}{\electronvolt},
and the structures are relaxed until all forces on each atom are below \qty{0.01}{\electronvolt\per\angstrom}.
For the 2D systems, a vacuum region of approximately \qty{22}{\angstrom} is included,
and a $\vec{k}$-point mesh of $20 \times 18 \times 1$ is used.
Convergence tests show that these settings yield accurate results,
as shown in figures~S1--S3 of the Supplementary Information.

For selected calculations,
we also use the screened hybrid exchange-correlation functional of Heyd, Scuseria, and Ernzerhof (HSE06),
which gives more accurate band-gap energies~\cite{sassnick2021electronic, krukau2006influence}.
Unless explicitly stated otherwise,
all figures show results obtained using the PBE functional.
For the density-of-states (DOS) calculations,
we use a $\Gamma$-centered $\vec{k}$-point mesh of $20 \times 28 \times 18$ for bulk o-\ce{B14},
and $20 \times 18 \times 1$ for the 2D systems.
The smearing at the Fermi level is calculated using the Gaussian method with a broadening value of \qty{0.1}{\electronvolt}.

The adsorption energy per atom, $E_{\mathrm{ad}}$, of hydrogen atoms on the surface is calculated as:

\begin{equation}
    E_{\mathrm{ad}} = \frac{E_{\mathrm{H/B}} - E_{\mathrm{B}} - E_{\mathrm{H}}}{n_{\mathrm{H}}}
\label{hydrogen_adsorption_energy}
\end{equation}

here,
$E_{\mathrm{H/B}}$ denotes the total energy of the hydrogen-adsorbed boron system,
$E_{\mathrm{B}}$ denotes the total energy of the pristine 2D boron structure,
$E_{\mathrm{H}}$ denotes the total energy of a hydrogen atom in the \ce{H2} molecule,
and $n_{\mathrm{H}}$ denotes the number of H atoms in the unit cell.

The cohesive energy per atom, $E_{\mathrm{c}}$, of the boron systems is calculated as:

\begin{equation}
    E_{\mathrm{c}}
    = \frac{E_{\mathrm{B\text{-}struc}} - n_{\mathrm{B}}E_{\mathrm{B\text{-}atom}}}{n_{\mathrm{B}}}
\label{cohesive_energy_per_atom}
\end{equation}

here,
$E_{\mathrm{B\text{-}struc}}$ denotes the total energy of the boron structure of interest,
$E_{\mathrm{B\text{-}atom}}$ denotes the total energy of a free boron atom calculated in a large supercell,
and $n_{\mathrm{B}}$ denotes the number of boron atoms in the unit cell.

The charge density difference, $\Delta \rho(\vec{r})$, for hydrogen adsorption on the surface is calculated as:

\begin{equation}
    \Delta \rho(\vec{r}) =
    \rho_{\mathrm{H/B}}(\vec{r})
    - \rho_{\mathrm{B}}(\vec{r})
    - \rho_{\mathrm{H}}(\vec{r})
\label{charge_density_difference_hydrogen_adsorption}
\end{equation}

here,
$\rho_{\mathrm{H/B}}(\vec{r})$ denotes the total charge density of the hydrogen-adsorbed system.
The terms $\rho_{\mathrm{B}}(\vec{r})$ and $\rho_{\mathrm{H}}(\vec{r})$ denote the charge densities of the clean boron structure and the adsorbed hydrogen atoms,
respectively.

The superconducting behaviour of o-\ce{B14} is investigated systematically in three structural phases,
namely bulk,
monolayer,
and hydrogen-terminated bilayer.
Calculations are carried out using the PBE formulation of the generalized gradient approximation~\cite{PBE-1996},
as provided in Quantum ESPRESSO~\cite{giannozzi2009quantum}.
Norm-conserving pseudopotentials and a kinetic energy cutoff of $45$ Ry are used~\cite{hamann2013optimized}.
For the 2D systems,
a vacuum spacing of \qty{15}{\angstrom} is added to suppress spurious periodic interactions.
Monkhorst--Pack $\vec{k}$-point meshes of $9 \times 12 \times 8$ for the bulk,
$10 \times 10 \times 1$ for the monolayer,
and $9 \times 9 \times 1$ for the bilayer are employed to sample the Brillouin zone~\cite{methfessel1989high}.
The phonon dispersions are calculated on $\vec{q}$-point grids of $3 \times 4 \times 3$,
$5 \times 5 \times 1$,
and $3 \times 3 \times 1$,
respectively~\cite{baroni2001phonons}.
Atomic positions and lattice parameters are relaxed until the residual forces are below $1\times10^{-5}$ Ry/bohr,
corresponding to \qty{2.6e-4}{\electronvolt\per\angstrom}.

Maximally localized Wannier functions are constructed in Wannier90 by initializing projections on B $s$ and $p$ orbitals~\cite{pizzi2020wannier90}.
Uniform $\vec{k}$-point meshes of $6 \times 8 \times 6$ for the bulk and $6 \times 6 \times 1$ for the monolayer and bilayer are used in Wannier interpolation.
Superconductivity is subsequently analyzed through EPW~\cite{giannozzi2009quantum,giannozzi2020quantum,giannozzi2017advanced},
which is used to evaluate the electron-phonon coupling strength and estimate the Allen--Dynes transition temperature $T_c$~\cite{margine2013anisotropic}.
The superconducting properties are also obtained by solving the fully anisotropic Migdal--Eliashberg equations~\cite{bardeen1955electron,giustino2017electron,margine2013anisotropic},
as implemented in EPW.
The Morel--Anderson pseudopotential $\mu^{*}$ is chosen to be $0.13$,
consistent with the typical semi-empirical range of $0.10$--$0.16$ used in Migdal--Eliashberg calculations for light-element and $sp$-bonded superconductors~\cite{allen-dynes},
and with the value adopted in the literature for boron-based 2D superconductors~\cite{zhao2019two}.
With this choice,
the calculated electron-phonon coupling strength $\lambda$ and $T_c$ indicate clear dimensionality effects,
as discussed below in the superconductivity section.

For the calculation of the optical properties within VASP,
we calculate the frequency-dependent dielectric tensor $\varepsilon_{\alpha\beta}(\omega)$.
The physical meaning of $\varepsilon_{\alpha\beta}(\omega)$,
its real and imaginary parts,
and the resulting optical observables are introduced in chapter~\ref{cha:fundamentals_b}.
In particular,
the dielectric response is described in section~\ref{sec:from_electronic_response_to_dielectric_function},
the Kramers--Kronig relation between the real and imaginary parts is discussed in section~\ref{sec:causality_and_kramers_kronig_relations},
and the absorption coefficient $\alpha_{\lambda}(\omega)$,
refractive index $n_{\lambda}(\omega)$,
extinction coefficient $k_{\lambda}(\omega)$,
reflectivity $R_{\lambda}(\omega)$,
and energy-loss spectrum $L_{\lambda}(\omega)$ are derived in section~\ref{sec:from_dielectric_function_to_optical_properties}.

For the present systems,
$\vec{k}$-point meshes of $20 \times 28 \times 18$ and $20 \times 18 \times 1$ are found to yield converged dielectric responses for the bulk and 2D systems,
respectively,
as shown in figure~S4.
A convergence check for the number of unoccupied bands included in the dielectric-function calculation is also performed.
Figure~S5 shows the result for bulk o-\ce{B14}.
The dielectric response is well converged with $128$ bands,
and this value is therefore used for all systems.

For 2D slab calculations,
the dielectric tensor contains the combined dielectric response of the 2D material and the vacuum region.
Therefore,
the magnitude of the calculated $\varepsilon_{1}(\omega)$ and $\varepsilon_{2}(\omega)$ depends linearly on the vacuum thickness~\cite{weng2018honeycomb, laturia2018dielectric}.
To obtain the intrinsic dielectric function of the 2D structures,
we renormalize the supercell dielectric function following Ref.~\cite{yang2021method}:

\begin{equation}
    \varepsilon_{2}^{\mathrm{2D}}(\omega)
    = \frac{d_{\mathrm{SC}}}{d_{\mathrm{2D}}}
    \varepsilon_{2}^{\mathrm{SC}}(\omega)
    \label{dielectric_vacuum_correction_imaginary}
\end{equation}

\begin{equation}
    \varepsilon_{1}^{\mathrm{2D}}(\omega) =
    1 + \frac{d_{\mathrm{SC}}}{d_{\mathrm{2D}}}
    \left[\varepsilon_{1}^{\mathrm{SC}}(\omega)-1\right]
    \label{dielectric_vacuum_correction_real}
\end{equation}

here, $\varepsilon_{i}^{\mathrm{2D}}(\omega)$ and $\varepsilon_{i}^{\mathrm{SC}}(\omega)$,
with $i=1,2$,
denote the dielectric functions of the 2D structure after removing the vacuum contribution and of the full supercell,
respectively. The quantity $d_{\mathrm{SC}}$ denotes the length of the supercell,
whereas $d_{\mathrm{2D}}$ denotes the effective thickness of the 2D structure.
The values of $d_{\mathrm{2D}}$ for the monolayer,
bilayer, and H-terminated bilayer are \qty{5.897}{\angstrom},
\qty{9.412}{\angstrom}, and \qty{11.209}{\angstrom}, respectively.
The definition of the thickness is shown in figure~S6.
In studies of 2D materials, omitting this correction can produce errors in the reported dielectric functions and therefore in the derived optical properties.

When interpreting the energy-loss spectra, we also consider that individual interband transitions may coexist with plasmons and form a broad background.
The coexistence of individual transitions and plasmons in the same energy range can couple the two excitation channels, leading to finite plasmon lifetimes as well as interband-transition contributions~\cite{li2017first}.

\section{Results and discussion}
\label{2_results}

\section{Conclusions}
\label{2_conclusions}

\section{Data availability statement}

\section{Supplementary information}
\label{2_supplementary}


