\chapter[Electronic and Optical Properties of 2D Heterostructure Bilayers]{Electronic and Optical Properties of 2D Heterostructure Bilayers of Graphene, Borophene and 2D Boron Carbides from First-Principles}
\label{cha:project_1}

\section{Abstract}

In the present work, 
the atomic, electronic, and optical properties of two-dimensional graphene, borophene, and boron carbide heterojunction bilayer systems (\ce{Graphene-BC3}, \ce{Graphene-Borophene}, and \ce{Graphene-B4C3}),
as well as their constituent monolayers are investigated on the basis of first-principles calculations using the HSE06 hybrid functional.
Our calculations show that while borophene is metallic,
both monolayer \ce{BC3} and \ce{B4C3} are indirect semiconductors,
with band-gaps of $1.822$ eV and $2.381$ eV as obtained using HSE06.
The \ce{Graphene-BC3} and \ce{Graphene-B4C3} bilayer heterojunction systems maintain the Dirac point of graphene at the K-point and are semi-metals,
while \ce{Graphene-Borophene} is metallic.
All bilayer heterostructure systems possess absorbance in the visible region where the resonance frequency and resonance absorption peak intensity vary between structures.
Remarkably, all heterojunctions support plasmons within the range $16.5-18.5$ eV, while \ce{Graphene-B4C3} and \ce{Graphene-Borophene} exhibit a $\pi$-type plasmon within the region 4-6 eV,
with the latter possessing an additional plasmon at the lower energy of $1.5-3$ eV.
The dielectric tensor for \ce{Graphene-B4C3} exhibits complex off-diagonal elements due to the lower P3 space group symmetry indicating it has anisotropic dielectric properties and could exhibit optically active (chiral) effects.
Our study shows that the two-dimensional heterostructures have desirable optical properties broadening their potential applications.

\newpage
\section{Introduction}

\ce{Graphene} is well-known to have excellent electrical conductivity,
desirable mechanical and thermal properties,
and high light transmittance in the visible light-infrared region.
It also has an extremely high quantum efficiency for light-matter interactions and is strongly optically nonlinear.
It has therefore found applications in e.g. electronics, energy storage, biomedical and other fields~\cite{janavika2023graphene}. 
Relative to conventional plasmonic materials, graphene possesses highly confined plasmons with much longer lifetimes.
Also, graphene plasmons are active in an extended wavelength range, namely, the mid-infrared and terahertz regime.
This wavelength range overlaps with the feature signals of most organic and biomolecules,
which has broadened graphene's applications towards plasmonic biological and chemical sensors~\cite{sakthivel2023} in addition to plasmonic waveguides,
and infrared photodetectors.
Moreover, the properties can be tuned by forming graphene-hybrid materials, finding applications in biosensors, chemical sensors, optical sensors, and other types of sensors~\cite{zhang2022}.

In other areas, however, the use of graphene has been limited due to its zero-value band gap.
One of the methods used to expand the application of graphene is to form heterostructures.
Stacking different two-dimensional (2D) materials together can form double-layer or even multi-layer artificial materials that are maintained by van der Waals interactions~\cite{novoselov20162d}.
Such materials are known as van der Waals heterojunctions.
A wide range of physical properties can be obtained by such stacking,
making van der Waals heterojunctions even more important than the 2D material itself.
There are vast numbers of 2D materials, and with forming heterostructures thereof,
a diverse range of electrical, photonic, plasmonic, mechanical, and chemical properties can be theoretically predicted through {\em ab initio} calculations and machine learning~\cite{thygesen2017,he2023}, and experimentally engineered~\cite{lin2023recent,elbanna20232d}.

Recently, the prospect of developing plasmonic devices employing 2D semiconductors has attracted considerable attention~\cite{agarwal2018plasmonics, qiu2018optical}. 
Plasmonic modes in each class of van der Waals semiconductors have their own peculiarities, along with potential technological capabilities.
Recently, graphene-like \ce{BC3} has been synthesized experimentally~\cite{yanagisawa2004phonon, yanagisawa2006phonon, tanaka2005novel} and is found to be semiconducting with an indirect bandgap.
Calculations have shown that nanostructures of graphene-like \ce{BC3} possess desirable absorbance in the visible region and by changing the size of the nanostructure,
the resonance peak position of the absorption spectrum can be effectively regulated~\cite{chen2019plasmon}.
Recent studies have also predicted a stable \ce{B4C3} monolayer,
which is even more stable than the previously synthesized BC3 monolayer~\cite{tian2019}.
Boron carbides are known for their exceptional hardness, 
thermal stability, and chemical inertness.
The emergence of 2D boron carbides adds a new dimension to the study of these materials,
offering novel properties and opportunities for technological innovation~\cite{luo2011predicting,fan2018two}.

Borophene is another 2D material of high current interest.
This monolayer allotrope of boron forms several structures and is characterized by its unique atomic structure and exceptional electronic properties,
which has opened new avenues for innovation in the areas of electronics,
energy storage, catalysis, plasmonics, superconductivity, and sensors~\cite{qiu2018optical}.
One particularly intriguing aspect of borophene's versatility lies in its ability to form heterojunctions—interfaces between different borophene structures or between borophene and other materials.
Borophene-graphene heterostructures have been realized~\cite{hou2021borophene} where they have shown good stability and ability to function as a humidity sensor,
exhibiting a sensitivity about 700 times higher than that of pristine graphene,
and 27 times higher than that of borophene.
Also, recently both lateral and vertical integration of borophene with graphene has been achieved~\cite{liu2019borophene},
as has the formation of bi-layer ($\alpha$-phase) borophene on the Ag(111) surface~\cite{liu2022}.

The synergistic combination of graphene's exceptional electrical conductivity,
mechanical strength, and flexibility with boron carbide's high hardness,
thermal stability, and chemical resistance, and with borophene's attractive electronic properties,
hold tremendous potential for a wide range of applications spanning from electronics and optoelectronics to energy storage and sensing.
As research in this field continues to advance,
the development of scalable synthesis methods, the elucidation of structure-property relationships,
and the exploration of novel applications will be key focal points.
To date, there is little known about the physical properties of graphene-based bilayer heterojunctions with borophene and boron-carbide materials.
This motivates the present work to gain an understanding of these heterojunctions, 
through the study of their electronic and optical properties on the basis of first-principles calculations.
Such understanding is crucial for their future technological application,
where the unique properties of these materials can be harnessed. 

The paper is organised as follows:
in Section 2 the calculation method is described,
followed by the Results and Discussion in Section 3.
Herein we describe the physical properties of the monolayers and heterostructures, including bandstructure, density of states, electron density difference distributions,
Schottky barriers and the optical properties.
This is followed by the  Conclusion in Section 4.

\section{Calculation Method}

% test test
