\chapter*{List of Publications}
\phantomsection
\addcontentsline{toc}{chapter}{List of Publications}

The following is a list of chapters in this thesis,
which are based on manuscripts that have been published,
accepted for publication,
or submitted to scientific journals.

Each chapter is a self-contained body of work and follows the usual structure of a scientific publication,
starting with an abstract and followed by an introduction, results, discussion, and conclusion.

\par\vspace{\baselineskip}

\subsubsection*{Chapter 4}

L. Niu, O. J. Conquest, C. Verdi, and C. Stampfl.
``Electronic and Optical Properties of 2D Heterostructure Bilayers of Graphene, Borophene and 2D Boron Carbides from First Principles''.
In: \textit{Nanomaterials} \textbf{14}.20 (2024), p. 1659.
\par\hspace*{1em}\texttt{doi:}\,\href{https://doi.org/10.3390/nano14201659}{\texttt{10.3390/nano14201659}}

\subsubsection*{Chapter 5}

L. Niu, O. J. Conquest, H. Ma, C. Verdi, and C. Stampfl.
``Superconductivity, Magnetism, and Directional Plasmonic Modes in Two-Dimensional Penta-Bipyramid Boron Allotropes''.
In: \textit{Physical Review Materials}, accepted for publication (2026).
\par\hspace*{1em}\texttt{doi:}\,\href{https://doi.org/10.1103/b2q9-fytt}{\texttt{10.1103/b2q9-fytt}}

\subsubsection*{Chapter 6}

L. Niu, O. J. Conquest, H. Ma, C. Verdi, and C. Stampfl.
``Emergent Topological and Optical Properties in Pristine and Hydrogen-Functionalized Two-Dimensional Beryllium Phases''.
Submitted to: \textit{Physical Review B}.

\par\vspace{\baselineskip}

\subsection*{Conferences that I attend during my PhD period}

\susubsection*{2025}

Feb. 4--Feb. 7, The 47th Annual Condensed Matter and Materials Meeting, Wagga 2025, University of Queensland, Brisbane, QLD 4072, Australia.

\susubsection*{2023}

Jul. 10--Jul. 12, NSW Student Computational Chemistry Meeting, The Sebel Harbourside Kiama, Kiama, NSW 2533, Australia.

\par\vspace{\baselineskip}

As supervisor for the candidature upon which this thesis is based,
I can confirm that the authorship attribution statements above are correct.

\par\vspace{\baselineskip}

Catherine Stampfl

\par\vspace{\baselineskip}

Oliver James Conquest

\par\vspace{\baselineskip}

Carla Verdi
